---
title: "Week 2 Data Cleaning Assignment"
subtitle: "Cleaning Water Cover, Land Cover, and Total Cover"
author: "Aakriti Poudel"
date: April 12, 2026
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    highlight-style: github
    code-fold: false
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| message: false

# Load packages
library(tidyverse)

# Folder paths
datadir_processed <- file.path("data", "processed")

# Input file (Snow_cover already cleaned in class)
snowsurvey_file <- "snow_cover.csv"
```

## Water cover data

::: {.callout-note title="Our task"}
We already cleaned `Snow_cover` in class. Now we clean `Water_cover` and
`Land_cover` using the same steps, then use the rule that all three covers
should sum to 100 to fill in missing values and fix `Total_cover`.
:::

### Data exploration

We start from the `snow_cover.csv` file produced at the end of the class
session. This file already has `Snow_cover` cleaned and stored as a number,
so we do not need to touch that column again.

```{r}
#| message: false

# Read the processed file from class
snowsurvey_csv <- read_csv(file.path(datadir_processed, snowsurvey_file))

glimpse(snowsurvey_csv)
```

Looking at the output above, `Water_cover` and `Land_cover` are both read
as character columns. According to the metadata file `01_ASDN_Readme.txt`,
these columns are defined as:

- `Water_cover`: percent cover of water
- `Land_cover`: percent cover of exposed land
- `Total_cover`: total sum to check the above percents; should always sum to 100

They should be numbers between 0 and 100, so something is preventing R from
treating them as numeric. Let's find out what:

```{r}
# Count all unique values in Water_cover
snowsurvey_csv %>%
  count(Water_cover)
```

Let's focus on just the non-numeric values:

```{r}
# Filter to non-numeric values only
snowsurvey_csv %>%
  count(Water_cover) %>%
  filter(is.na(as.numeric(Water_cover)))
```

We have 767 rows with non-numeric values across `.`, `-`, `n/a`, and `unk`.
That is a meaningful number of rows, so we need to handle each one carefully
before we can convert the column to numeric.

### Data cleaning

We found the same problematic values as in `Snow_cover` during class. The
metadata does not define any of these symbols as valid values for cover
columns, so our general decision is to treat them all as missing (NA).
Below we look at each one in detail before replacing it.

#### `.` values

```{r}
#| eval: false
# Inspect rows with a dot
snowsurvey_csv %>%
  filter(Water_cover == ".") %>%
  View()
```

There are 575 rows with a dot in `Water_cover`. Looking at those rows, the
`Snow_cover` column was already set to NA during class cleaning, and
`Land_cover` also has a dot in the same rows. This pattern suggests the
entire observation was not recorded on that day, not just `Water_cover`.
The metadata (`01_ASDN_Readme.txt`) defines every column in the Snow_survey
table but makes no mention of a dot as a valid value for any cover column.
Because we have no definition for it and no way to recover the true value,
we treat it as missing and replace it with NA.

```{r}
# Replace dot with NA -- not defined in metadata
snowsurvey_fixed <- snowsurvey_csv %>%
  mutate(Water_cover = ifelse(Water_cover == ".", NA, Water_cover))
```

#### `-` values

Is the problem similar with `-`?

```{r}
#| eval: false
# Inspect rows with a dash
snowsurvey_csv %>%
  filter(Water_cover == "-") %>%
  View()
```

There are 10 rows with a dash. In those rows, `Land_cover` also has a dash
and `Snow_cover` is NA. Interestingly, the metadata does define `-` in one
place: in the color band codes table it means "no bands on this part of the
leg." However, that definition belongs to a completely different data table
and has nothing to do with cover percentages. For the Snow_survey table, `-`
is not defined anywhere. It looks like it was used as a placeholder when no
value was recorded, the same way some people write a dash in a paper form to
mean "nothing here." We replace it with NA.

```{r}
# Replace dash with NA -- not a valid code in metadata
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Water_cover = ifelse(Water_cover == "-", NA, Water_cover))
```

#### `n/a` values

`n/a` seems pretty clear, but let's have a look first:

```{r}
#| eval: false
# Inspect rows with n/a
snowsurvey_csv %>%
  filter(Water_cover == "n/a") %>%
  View()
```

Same pattern as the dot and dash rows. The metadata does not define `n/a`
as a valid entry. It almost certainly means "not available", which is just
another way of saying the value is missing. We replace it with NA.

```{r}
# Replace n/a with NA -- means not available
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Water_cover = ifelse(Water_cover == "n/a", NA, Water_cover))
```

#### `unk` values

`unk` is probably short for unknown:

```{r}
#| eval: false
# Inspect the row with unk
snowsurvey_csv %>%
  filter(Water_cover == "unk") %>%
  View()
```

There is only 1 row with `unk`. Looking at it, `Snow_cover` is NA and
`Land_cover` is also `unk`. The metadata does not mention this code for
cover columns. Since we do not know the actual value and have no way to
recover it, we replace it with NA.

```{r}
# Replace unk with NA -- unknown, no value to work with
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Water_cover = ifelse(Water_cover == "unk", NA, Water_cover))
```

**Note about `<1`:** In class, `Snow_cover` had a `<1` value. We replaced
it with `0` because a snow cover of less than 1% is effectively 0 at the
whole-number scale we are working at. `Water_cover` does not have any `<1`
values, so we skip that step here.

### Going numeric

Now let's check that only NAs are left as non-numeric values:

```{r}
# Confirm only NAs remain before converting
snowsurvey_fixed %>%
  count(Water_cover) %>%
  filter(is.na(as.numeric(Water_cover)))
```

Only NAs remain. We can safely convert the column to numeric now.

```{r}
# Convert to numeric
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Water_cover = as.numeric(Water_cover))
```

We have a numeric column now 🎉

### Further cleaning

We are dealing with percentages, so all values must be between 0 and 100.
Let's check:

```{r}
# Check for values above 100
snowsurvey_fixed %>%
  filter(Water_cover > 100)
```

There is one row where `Water_cover` is 353. That is clearly not a valid
percentage. Looking at the full row, `Snow_cover` is 45 and `Land_cover`
is -298. If you add those three together: 45 + 353 + (-298) = 100, which
matches the recorded `Total_cover`. This tells us the numbers were probably
swapped or mistyped during data entry. The true split between water and land
could be almost anything, so we cannot make a reasonable guess. We set
`Water_cover` to NA. We will also deal with the `-298` in `Land_cover`
when we get to that column.

```{r}
# Set values above 100 to NA
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Water_cover = ifelse(Water_cover > 100, NA, Water_cover))
```

Let's check for negative values:

```{r}
# Check for negative values
snowsurvey_fixed %>%
  filter(Water_cover < 0)
```

No negative values in `Water_cover` ✅

Decimal points don't really make sense for a percent cover field survey.
The metadata defines `Water_cover` as "percent cover of water" but does not
say whether values should be whole numbers or can have decimals. Looking at
the unique values, we can see some entries like 0.25, 1.75, etc. Our
assumption is that these are rounding artifacts from data entry rather than
intentional precision, because a field observer estimating percent cover by
eye would not realistically record values like 1.75%. We round to whole
numbers to be consistent with how most of the data are recorded.

```{r}
# Check for decimal values before rounding
unique(snowsurvey_fixed$Water_cover)
```

```{r}
# Round to whole numbers
snowsurvey_fixed$Water_cover <- round(snowsurvey_fixed$Water_cover)
# Confirm no decimals remain
unique(snowsurvey_fixed$Water_cover)
```

---

## Land cover data

### Data exploration

Now let's look at `Land_cover`:

```{r}
# Filter to non-numeric values only
snowsurvey_csv %>%
  count(Land_cover) %>%
  filter(is.na(as.numeric(Land_cover)))
```

We have 772 rows with non-numeric values. The same four values appear:
`.`, `-`, `n/a`, and `unk`. We apply the same decisions as we did for
`Water_cover` since the reasoning is identical and the metadata gives no
special meaning to any of these symbols in cover columns.

### Data cleaning

#### `.` values

```{r}
#| eval: false
# Inspect rows with a dot
snowsurvey_csv %>%
  filter(Land_cover == ".") %>%
  View()
```

There are 585 rows with a dot in `Land_cover`. As with `Water_cover`, the
dot rows also have dots or NAs in the other cover columns. No value was
recorded for the whole survey on those days. We replace with NA.

```{r}
# Replace dot with NA -- not defined in metadata
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Land_cover = ifelse(Land_cover == ".", NA, Land_cover))
```

#### `-` values

```{r}
#| eval: false
# Inspect rows with a dash
snowsurvey_csv %>%
  filter(Land_cover == "-") %>%
  View()
```

Same pattern as `Water_cover`. Ten rows, all with dashes or NAs across the
cover columns. We replace with NA.

```{r}
# Replace dash with NA -- not a valid code in metadata
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Land_cover = ifelse(Land_cover == "-", NA, Land_cover))
```

#### `n/a` values

```{r}
#| eval: false
# Inspect rows with n/a
snowsurvey_csv %>%
  filter(Land_cover == "n/a") %>%
  View()
```

Same pattern. We replace with NA.

```{r}
# Replace n/a with NA -- means not available
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Land_cover = ifelse(Land_cover == "n/a", NA, Land_cover))
```

#### `unk` values

```{r}
#| eval: false
# Inspect the row with unk
snowsurvey_csv %>%
  filter(Land_cover == "unk") %>%
  View()
```

One row, same row as the `unk` in `Water_cover`. We replace with NA.

```{r}
# Replace unk with NA -- unknown, no value to work with
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Land_cover = ifelse(Land_cover == "unk", NA, Land_cover))
```

**Note about `<1`:** Just like `Water_cover`, `Land_cover` has no `<1`
values, so we do not need that step here either.

### Going numeric

Check that only NAs are left as non-numeric:

```{r}
# Confirm only NAs remain before converting
snowsurvey_fixed %>%
  count(Land_cover) %>%
  filter(is.na(as.numeric(Land_cover)))
```

Good. We can convert:

```{r}
# Convert to numeric
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Land_cover = as.numeric(Land_cover))
```

We have a numeric column now 🎉

### Further cleaning

Check for values above 100:

```{r}
# Check for values above 100
snowsurvey_fixed %>%
  filter(Land_cover > 100)
```

No values above 100 in `Land_cover` ✅

Let's check for negative values:

```{r}
# Check for negative values
snowsurvey_fixed %>%
  filter(Land_cover < 0)
```

There are two negative rows. Percent cover of exposed land cannot be
negative -- that has no physical meaning in the real world.

- The row with `-298` is the same row we saw earlier with `Water_cover = 353`.
  As discussed above, both values are clearly wrong due to a data entry
  error. We already set `Water_cover` to NA. We set `Land_cover` to NA here
  for the same reason.

- The row with `-100` has `Snow_cover = NA`, `Water_cover = 0`, and
  `Total_cover = 100`. If we trusted the total and tried to work backward,
  Snow + Land = 100, and with Land = -100 that would require Snow = 200,
  which is also impossible. So there is no internally consistent way to
  recover this row. We set `Land_cover` to NA.

In both cases, setting to NA is the safest choice because making up a value
would introduce error into any analysis that uses this data.

```{r}
# Set negative values to NA
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Land_cover = ifelse(Land_cover < 0, NA, Land_cover))
```

Same assumption as `Water_cover`: the metadata does not say values must be
whole numbers, but decimal values like 1.5 or 2.75 are unlikely to reflect
real field precision. We round to whole integers.

```{r}
# Check for decimal values before rounding
unique(snowsurvey_fixed$Land_cover)
```

```{r}
# Round to whole numbers
snowsurvey_fixed$Land_cover <- round(snowsurvey_fixed$Land_cover)
# Confirm no decimals remain
unique(snowsurvey_fixed$Land_cover)
```

No negative values detected ✅

---

## Inferring missing values and fixing `Total_cover`

### Why we can fill in some missing values

The metadata says `Total_cover` is the sum of `Snow_cover`, `Water_cover`,
and `Land_cover`, and it should always equal 100. This makes sense because
the three cover types are meant to account for 100% of the ground surface.

This gives us a useful way to recover some missing values. If a row has a
valid number in two of the three cover columns and NA in the third, we can
work out the missing value:

`missing value = 100 - (sum of the other two)`

For example, if `Snow_cover = 80` and `Water_cover = 10` and `Land_cover`
is NA, then `Land_cover` must be `100 - 80 - 10 = 10`.

We apply this rule only when exactly one column is missing. If two or all
three are missing, we do not have enough information and we leave them as NA.
We also check that the inferred value falls between 0 and 100. If it does
not, the two known values are already inconsistent and we should not trust
the inferred one.

One important assumption here: we trust that the two known values are
correct. If they were themselves recorded with errors, the inferred value
would also be wrong. We accept this risk because we have no other
information to go on.

Across the whole dataset, this rule lets us recover 96 missing values.

```{r}
# Infer missing value when exactly one cover column is NA
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(
    # Count NAs across the three cover columns
    na_count = is.na(Snow_cover) + is.na(Water_cover) + is.na(Land_cover),

    # Sum of the non-NA cover values
    known_sum = rowSums(cbind(
      ifelse(is.na(Snow_cover),  0, Snow_cover),
      ifelse(is.na(Water_cover), 0, Water_cover),
      ifelse(is.na(Land_cover),  0, Land_cover)
    )),

    # Fill Snow_cover if it is the only missing value
    Snow_cover = ifelse(
      na_count == 1 & is.na(Snow_cover) &
        (100 - known_sum) >= 0 & (100 - known_sum) <= 100,
      100 - known_sum,
      Snow_cover
    ),

    # Fill Water_cover if it is the only missing value
    Water_cover = ifelse(
      na_count == 1 & is.na(Water_cover) &
        (100 - known_sum) >= 0 & (100 - known_sum) <= 100,
      100 - known_sum,
      Water_cover
    ),

    # Fill Land_cover if it is the only missing value
    Land_cover = ifelse(
      na_count == 1 & is.na(Land_cover) &
        (100 - known_sum) >= 0 & (100 - known_sum) <= 100,
      100 - known_sum,
      Land_cover
    )
  ) %>%
  # Drop helper columns
  select(-na_count, -known_sum)
```

### Recompute `Total_cover`

Now we recompute `Total_cover` directly from the three cleaned cover columns.
We do this for two reasons:

1. The original `Total_cover` column had its own non-numeric values like `.`,
   `-`, and `n/a`. Recomputing fixes those automatically.
2. After our cleaning, some rows now have different values for the three
   cover columns than before (the 96 recovered values, plus the rows we set
   to NA). Recomputing makes sure `Total_cover` reflects the current state
   of the data, not the original raw state.

If a row still has any NA in the three cover columns after recovery, its
`Total_cover` will also be NA. That is the correct behavior because we
cannot give a reliable total if we are missing one of the parts.

```{r}
# Recompute Total_cover from the three cleaned columns
snowsurvey_fixed <- snowsurvey_fixed %>%
  mutate(Total_cover = Snow_cover + Water_cover + Land_cover)
```

Let's check the result:

```{r}
# Check the distribution of Total_cover values
snowsurvey_fixed %>%
  count(Total_cover) %>%
  arrange(desc(n))
```

Most rows have a total of 100, which is what the metadata says we should
expect. Some rows have totals other than 100. The metadata states the total
"should always sum to 100," but this appears to be an expectation rather
than something that was enforced when the data were collected. We cannot
correct these rows without knowing what the true values should be. For
example, 1,281 rows have all three covers recorded as 0, giving a total of
0. These appear to be real field observations -- possibly from early in the
season before any snow, water, or land was visible -- where the observer
recorded 0 for each cover type but did not update the total. We leave them
as is and accept that some rows will not meet the metadata expectation.

Let's write the cleaned file to a csv:

```{r}
# Create output folder if needed
dir.create(datadir_processed, showWarnings = FALSE)

# Save the cleaned dataset
write_csv(snowsurvey_fixed, file.path(datadir_processed, "all_cover_fixed_AakritiPoudel.csv"))
```
